BIP-Predictor

### 2025-05-08 Sweave/BIP_predictor.Rnw
up

  # We can now plot target and direct forecast: for h>2 the predictor comes close to a flat line centered at zero
  mplot<-cbind(forward_shifted_BIP,direct_forecast)
  rownames(mplot)<-rownames(x_mat_wc)
  colnames(mplot)<-c("BIP","Direct forecast")
  colo<-c("black","red")
  main_title<-paste("BIP shifted  by ",h," quarters",sep="")
  plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
  for (i in 1:ncol(mplot))
  {
    lines(mplot[,i],col=colo[i])
    mtext(colnames(mplot)[i],col=colo[i],line=-i)
  }
  abline(h=0)
  axis(1,at=c(1,12*1:(nrow(mplot)/12)),labels=rownames(mplot)[c(1,12*1:(nrow(mplot)/12))])
  axis(2)
  box()
}
dev.off()

@@

  # We can now plot target and direct forecast: for h>2 the predictor comes close to a flat line centered at zero
  mplot<-cbind(forward_shifted_BIP,direct_forecast)
  rownames(mplot)<-rownames(x_mat_wc)
  mplot<-mplot[which(rownames(mplot)>2008&rownames(mplot)<2011),]
  colnames(mplot)<-c("BIP","Direct forecast")
  colo<-c("black","red")
  main_title<-paste("BIP shifted forward by ",h," quarters",sep="")
  plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
  for (i in 1:ncol(mplot))
  {
    lines(mplot[,i],col=colo[i])
    mtext(colnames(mplot)[i],col=colo[i],line=-i)
  }
  abline(h=0)
  axis(1,at=1:nrow(mplot),labels=rownames(mplot))
  axis(2)
  box()
}
dev.off()
@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/direct_wc_all.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=4in, width=5in]{", file, "}\n",sep = "")
cat("\\caption{Direct forecasts (red lines) based on all indicators and full sample information, without the pandemic: BIP (black line) shifted by zero one, two and three quarters.", sep = "")
cat("\\label{direct_wc}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/direct_wc_financial_crisis.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=4in, width=5in]{", file, "}\n",sep = "")
cat("\\caption{Direct forecast  across the financial crisis:  regression based on all indicators and full sample information.", sep = "")
cat("\\label{direct_wc_financial_crisis}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
The results indicate that as $h$ increases, the peaks and dips of the direct forecasts become increasingly right-shifted (delayed) relative to the target. Additionally, the F-statistics precipitously decline from $F=\Sexpr{round(f_vec_unfiltered[3],2)}$, for $h=2$, to $F=\Sexpr{round(f_vec_unfiltered[4],2)}$, for $h=3$, see Table \eqref{f_stat}, suggesting that the predictor becomes statistically insignificant for larger forecast horizons exceeding two quarters (link to Heinisch and Scheufele).

@@

file = "./Figures/hp_160.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 6, height = 6)
par(mfrow=c(2,2))
hp_two<-c(HP_obj$hp_mse[L:2],HP_obj$hp_mse)
ts.plot(hp_two,main=paste("Two-sided HP(",lambda_HP,")",sep=""),xlab="",ylab="")
hp_c<-HP_obj$hp_trend
ts.plot(hp_c,main="One-sided (concurrent) HP-C",xlab="",ylab="")

@@

@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/hp_160.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=3in, width=4.5in]{", file, "}\n",sep = "")
cat("\\caption{Two-sided HP-160) (top left) and one-sided (concurrent) HP-C filters (top right) with corresponding amplitude functions (bottom).", sep = "")
cat("\\label{hp_160}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@

@@

  colnames(mplot)<-c("BIP","Direct forecast")
  colo<-c("black","red")
  main_title<-paste("BIP shifted forward by ",h," quarters",sep="")
  plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
  for (i in 1:ncol(mplot))
  {
    lines(mplot[,i],col=colo[i])
    mtext(colnames(mplot)[i],col=colo[i],line=-i)
  }
  abline(h=0)
  axis(1,at=1:nrow(mplot),labels=rownames(mplot))
  axis(2)
  box()
}
colnames(t_mat_hp)<-select_vec_multi
rownames(t_mat_hp)<-paste("h=",0:3)
dev.off()

@@

  colo<-c("black","red","blue")
  main_title<-paste("GDP shifted forward by ",h," quarters",sep="")
  plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
  for (i in 1:ncol(mplot))
  {
    lines(mplot[,i],col=colo[i])
    if (h==0)
    mtext(colnames(mplot)[i],col=colo[i],line=-i)
  }
  abline(h=0)
  axis(1,at=1:nrow(mplot),labels=rownames(mplot))
  axis(2)
  box()
}

dev.off()
@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/direct_hp_forecasts.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=4in, width=5in]{", file, "}\n",sep = "")
cat("\\caption{Direct HP forecasts: entire data set with exclusion of the pandemic.", sep = "")
cat("\\label{direct_hp_forecasts}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/direct_hp_forecasts_financial_crisis.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=5in, width=6in]{", file, "}\n",sep = "")
cat("\\caption{Shifted GDP (black), direct HP forecast (red) and classic direct forecast (blue) over the curse of the financial crisis:  regression based on all indicators and full sample information (without pandemic).", sep = "")
cat("\\label{direct_hp_forecasts_financial_crisis}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
Moreover, the F-statistics are larger than for the direct forecasts, see Table \eqref{f_stat}, but the predictor remains statistically insignificant at forecast horizons exceeding two quarters (even though the evaluation is conducted in-sample and based on all available indicators). 
<<label=ats_mba_2,echo=FALSE,results=tex>>=

@@

file = "./Figures/bk_gammak.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 6, height = 6)

par(mfrow=c(1,2))
mplot<-gammak
colo<-rainbow(ncol(mplot))
ts.plot(mplot,col=colo,main="M-MSE",xlab="")
for (i in 1:ncol(mplot))
  mtext(colnames(mplot)[i],col=colo[i],line=-i)
mplot<-bk
colo<-rainbow(ncol(mplot))
ts.plot(mplot,col=colo,main="M-SSA",xlab="")
for (i in 1:ncol(mplot))
  mtext(colnames(mplot)[i],col=colo[i],line=-i)
#mplot<-bk_HP_HP_C
#colo<-rainbow(ncol(mplot))
#ts.plot(mplot,col=colo,main="M-SSA larger HT")
#for (i in 1:ncol(mplot))
#  mtext(colnames(mplot)[i],col=colo[i],line=-i)

@@

file = "./Figures/mssa_msse_now.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 6, height = 6)

hp_c_out<-filter(x_mat[,select],hp_c,side=1)
par(mfrow=c(1,1))
#mplot<-scale(cbind(target_shifted_mat,hp_c_out,mssa[,select],mssa_HT_HP_C[,select],mmse[,select]))
#dim(mplot)
#colnames(mplot)<-c("Target: HP-BIP","HP-C","M-SSA smaller HT","M-SSA larger HT","M-MSE")
mplot<-scale(cbind(target_shifted_mat,mssa[,select],mmse[,select]))
dim(mplot)
colnames(mplot)<-c("Target: HP-BIP","M-SSA","M-MSE")

rownames(mplot)<-rownames(x_mat)
colo<-c("black","violet","blue","orange","green")
colo<-c("black","blue","green")
main_title<-"Nowcasting HP-BIP"
plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],lwd=1,ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
mtext(colnames(mplot)[1],col=colo[1],line=-1)
for (i in 1:ncol(mplot))
{
  lines(mplot[,i],col=colo[i],lwd=1,lty=1)
  mtext(colnames(mplot)[i],col=colo[i],line=-i)
}
abline(h=0)
abline(v=which(rownames(mplot)>date_to_fit)[1],lty=2,lwd=2)
axis(1,at=c(1,4*1:(nrow(mplot)/4)),labels=rownames(mplot)[c(1,4*1:(nrow(mplot)/4))])
axis(2)
box()

@@

mat_cor_ht<-rbind(c(cor(na.exclude(mplot[,c("Target: HP-BIP","M-SSA")]))[1,2],
cor(na.exclude(mplot[,c("Target: HP-BIP","M-MSE")]))[1,2]),
c(compute_empirical_ht_func(na.exclude(mssa[,select]))$empirical_ht,
compute_empirical_ht_func(na.exclude(mmse[,select]))$empirical_ht))

@@

Figure \ref{bk_gammak} displays the coefficients of the multivariate nowcasts, where increased smoothing achieved through the M-SSA method manifests as a distinct pattern in the filter weights, notably influencing the rate of decay across increasing lags. The resulting filtered series, presented in Figure \ref{mssa_msse_now}, are standardized to enable direct comparisons. The M-SSA nowcast adheres to the specified constraint, exhibiting approximately one-third fewer zero-crossings than the M-MSE, confirming the effectiveness of the smoothness control imposed during the design. This is further corroborated by Table \ref{corhtnow}, which reports empirical holding times and target correlations. Increased smoothness achieved by the M-SSA comes at the cost of a slight reduction in tracking accuracy, as indicated by a marginally lower target correlation. This highlights the inherent trade-off—often referred to as the prediction dilemma—between smoothness (zero-crossing rate) and predictive accuracy (target correlation) within the M-SSA criterion.
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/bk_gammak.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=3in, width=4.5in]{", file, "}\n",sep = "")
cat("\\caption{Multivariate M-MSE (left) and M-SSA  (right) nowcast filters.", sep = "")
cat("\\label{bk_gammak}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/mssa_msse_now.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=3in, width=4.5in]{", file, "}\n",sep = "")
cat("\\caption{Two-sided HP(160) applied to BIP (target, black) and nowcasts:  M-SSA (blue) and  M-MSE (green). The dashed vertical line delimits in- and out-of-sample spans. The first L=31 observations are lost due to initialization and the two-sided filter does not extend to the sample end.", sep = "")
cat("\\label{mssa_msse_now}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
<<label=ats_mba_2,echo=FALSE,results=tex>>=
mat1<-round(mat_cor_ht,3)
#latex(cor_vec, dec = 1, , caption = "Example of using latex to create table",
#center = "centering", file = "", floating = FALSE)
xtable(mat1, dec = 1,digits=rep(2,dim(mat1)[2]+1),
paste("Sample target correlations and HTs of M-SSA and M-MSE nowcasts.  "),
label=paste("corhtnow",sep=""),
center = "centering", file = "", floating = FALSE)
@
The results validate that the smoothing constraint incorporated into the multivariate extension of Criterion \eqref{critssa} effectively manages the zero-crossings of the predictor: sample estimates of the holding times (HTs) align with the expected values derived from Equation \eqref{ht}\footnote{Effective convergence of sample estimates to their theoretical expected values —assuming the true data-generating process is known— can be validated using the M-SSA package. This validation is typically performed through analyses of very long samples of simulated (multivariate) data, ensuring the consistency of the estimation procedures under idealized conditions.}. An increase of $50\%$ in the HT relative to the M-MSE predictor has only a modest effect on the target correlation, suggesting that this tradeoff could be beneficial in our application. Going forward, we now omit the M-MSE, as it is a special case of the more general M-SSA framework.

@@

file = "./Figures/bk_h.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 6, height = 6)

par(mfrow=c(1,2))
# h=0, see above
mplot<-bk
colo<-rainbow(ncol(mplot))
ts.plot(mplot,col=colo,main="M-SSA: h=0")
for (i in 1:ncol(mplot))
  mtext(colnames(mplot)[i],col=colo[i],line=-i)

# h=4
bk<-matrix(mssa_indicator_obj$bk_x_array[select,select,,"h=4"],nrow=L)
colnames(bk)<-colnames(gammak)<-select_vec_multi
mplot<-bk
colo<-rainbow(ncol(mplot))
ts.plot(mplot,col=colo,main="M-SSA: h=4")
for (i in 1:ncol(mplot))
  mtext(colnames(mplot)[i],col=colo[i],line=-i)

@@

file = "./Figures/multivar_vs_univar.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 6, height = 6)

# Select BIP
i<-1
# HP-C Financial crisis
mplot<-scale(hp_c_array[i,,])
colnames(mplot)<-paste("HP-",colnames(x_mat)[i],": h=",h_vec,sep="")
par(mfrow=c(2,2))
colo<-c(rainbow(ncol(mplot)))
mplot<-mplot[which(rownames(mplot)>2007&rownames(mplot)<2011),]
main_title<-paste("HP-C targeting HP-",colnames(x_mat)[i],sep="")
main_title<-"HP-C: financial crisis"
plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
mtext(colnames(mplot)[1],col=colo[1],line=-1)
for (j in 1:ncol(mplot))
{
  lines(mplot[,j],col=colo[j],lwd=1,lty=1)
  mtext(colnames(mplot)[j],col=colo[j],line=-j)
}
abline(h=0)
axis(1,at=1:nrow(mplot),labels=rownames(mplot))
axis(2)
box()

# HP-C Pandemic
mplot<-scale(hp_c_array[i,,])
colnames(mplot)<-paste("HP-",colnames(x_mat)[i],": h=",h_vec,sep="")
colo<-c(rainbow(ncol(mplot)))
mplot<-mplot[which(rownames(mplot)>2017&rownames(mplot)<2022),]
main_title<-paste("HP-C targeting HP-",colnames(x_mat)[i],sep="")
main_title<-"HP-C: pandemic"
plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
mtext(colnames(mplot)[1],col=colo[1],line=-1)
for (j in 1:ncol(mplot))
{
  lines(mplot[,j],col=colo[j],lwd=1,lty=1)
  mtext(colnames(mplot)[j],col=colo[j],line=-j)
}
abline(h=0)
axis(1,at=1:nrow(mplot),labels=rownames(mplot))
axis(2)
box()

@@

colo<-c(rainbow(ncol(mplot)))
main_title<-paste("M-SSA targeting HP-",colnames(x_mat)[i],sep="")
main_title<-"MSSA: financial crisis"
plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
mtext(colnames(mplot)[1],col=colo[1],line=-1)
for (j in 1:ncol(mplot))
{
  lines(mplot[,j],col=colo[j],lwd=1,lty=1)
  mtext(colnames(mplot)[j],col=colo[j],line=-j)
}
abline(h=0)
axis(1,at=1:nrow(mplot),labels=rownames(mplot))
axis(2)
box()

@@

# Plot of M-SSA and HP-C, only financial crisis
file = "./Figures/mssa_hpc_financial_crisis.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 6, height = 6)

@@

file = "./Figures/revisions3.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 6, height = 6)
par(mfrow=c(1,1))

>>>>

  # We can now plot target and direct forecast: for h>2 the predictor comes close to a flat line centered at zero
  mplot<-cbind(forward_shifted_BIP,direct_forecast)
  rownames(mplot)<-rownames(x_mat_wc)
  colnames(mplot)<-c("GDP","Direct forecast")
  colo<-c("black","red")
  main_title<-paste("GDP shifted  by ",h," quarters",sep="")
  plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
  for (i in 1:ncol(mplot))
  {
    lines(mplot[,i],col=colo[i])
    if (h==0)
    mtext(colnames(mplot)[i],col=colo[i],line=-i)
  }
  abline(h=0)
  axis(1,at=c(1,12*1:(nrow(mplot)/12)),labels=rownames(mplot)[c(1,12*1:(nrow(mplot)/12))])
  axis(2)
  box()
}
dev.off()

@@

  # We can now plot target and direct forecast: for h>2 the predictor comes close to a flat line centered at zero
  mplot<-cbind(forward_shifted_BIP,direct_forecast)
  rownames(mplot)<-rownames(x_mat_wc)
  mplot<-mplot[which(rownames(mplot)>2008&rownames(mplot)<2011),]
  colnames(mplot)<-c("GDP","Direct forecast")
  colo<-c("black","red")
  main_title<-paste("GDP shifted forward by ",h," quarters",sep="")
  plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
  for (i in 1:ncol(mplot))
  {
    lines(mplot[,i],col=colo[i])
    if (h==0)
    mtext(colnames(mplot)[i],col=colo[i],line=-i)
  }
  abline(h=0)
  axis(1,at=1:nrow(mplot),labels=rownames(mplot))
  axis(2)
  box()
}
dev.off()
@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/direct_wc_all.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=5in, width=6in]{", file, "}\n",sep = "")
cat("\\caption{Direct forecasts (red lines) based on all indicators and full sample information, without the pandemic: GDP (black line) shifted by zero one, two and three quarters.", sep = "")
cat("\\label{direct_wc}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/direct_wc_financial_crisis.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=5in, width=6in]{", file, "}\n",sep = "")
cat("\\caption{Direct forecast  across the financial crisis:  regression based on all indicators and full sample information.", sep = "")
cat("\\label{direct_wc_financial_crisis}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
The results indicate that as $h$ increases, the peaks and dips of the direct forecasts become increasingly right-shifted (delayed) relative to the target. Additionally, the F-statistics precipitously decline from $F=\Sexpr{round(f_vec_unfiltered[3],2)}$, for $h=2$, to $F=\Sexpr{round(f_vec_unfiltered[4],2)}$, for $h=3$, see Table \eqref{f_stat}, suggesting that the predictor becomes statistically insignificant for larger forecast horizons exceeding two quarters (link to Heinisch and Scheufele).

@@

file = "./Figures/hp_160.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7.5, height = 6)
par(mfrow=c(2,2))
hp_two<-c(HP_obj$hp_mse[L:2],HP_obj$hp_mse)
ts.plot(hp_two,main=paste("Two-sided HP(",lambda_HP,")",sep=""),xlab="",ylab="")
hp_c<-HP_obj$hp_trend
ts.plot(hp_c,main="One-sided (concurrent) HP-C",xlab="",ylab="")

@@

@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/hp_160.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=5in, width=6in]{", file, "}\n",sep = "")
cat("\\caption{Two-sided HP-160) (top left) and one-sided (concurrent) HP-C filters (top right) with corresponding amplitude functions (bottom).", sep = "")
cat("\\label{hp_160}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@

@@

  colnames(mplot)<-c("GDP","Direct forecast")
  colo<-c("black","blue")
  main_title<-paste("GDP shifted forward by ",h," quarters",sep="")
  plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
  for (i in 1:ncol(mplot))
  {
    lines(mplot[,i],col=colo[i])
    if (h==0)
    mtext(colnames(mplot)[i],col=colo[i],line=-i)
  }
  abline(h=0)
  axis(1,at=1:nrow(mplot),labels=rownames(mplot))
  axis(2)
  box()
}
colnames(t_mat_hp)<-select_vec_multi
rownames(t_mat_hp)<-paste("h=",0:3)
dev.off()

@@

  colo<-c("black","blue","red")
  main_title<-paste("GDP shifted forward by ",h," quarters",sep="")
  plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
  for (i in 1:ncol(mplot))
  {
    lines(mplot[,i],col=colo[i])
    if (h==0)
    mtext(colnames(mplot)[i],col=colo[i],line=-i)
  }
  abline(h=0)
  axis(1,at=1:nrow(mplot),labels=rownames(mplot))
  axis(2)
  box()
}

dev.off()
@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/direct_hp_forecasts.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=5in, width=6in]{", file, "}\n",sep = "")
cat("\\caption{Direct HP forecasts: entire data set with exclusion of the pandemic.", sep = "")
cat("\\label{direct_hp_forecasts}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/direct_hp_forecasts_financial_crisis.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=5in, width=6in]{", file, "}\n",sep = "")
cat("\\caption{Shifted GDP (black), direct HP forecast (blue) and classic direct forecast (red) over the curse of the financial crisis:  regression based on all indicators and full sample information (without pandemic).", sep = "")
cat("\\label{direct_hp_forecasts_financial_crisis}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
Moreover, the F-statistics are larger than for the direct forecasts, see Table \eqref{f_stat}, but the predictor remains statistically insignificant at forecast horizons exceeding two quarters (even though the evaluation is conducted in-sample and based on all available indicators). 
<<label=ats_mba_2,echo=FALSE,results=tex>>=

@@

file = "./Figures/bk_gammak.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7, height = 7)

par(mfrow=c(1,2))
mplot<-gammak
colnames(mplot)[1]<-"GDP"
colo<-rainbow(ncol(mplot))
ts.plot(mplot,col=colo,main="M-MSE",xlab="")
for (i in 1:ncol(mplot))
  mtext(colnames(mplot)[i],col=colo[i],line=-i)
mplot<-bk
colnames(mplot)[1]<-"GDP"

colo<-rainbow(ncol(mplot))
ts.plot(mplot,col=colo,main="M-SSA",xlab="")
#for (i in 1:ncol(mplot))
#  mtext(colnames(mplot)[i],col=colo[i],line=-i)
#mplot<-bk_HP_HP_C
#colo<-rainbow(ncol(mplot))
#ts.plot(mplot,col=colo,main="M-SSA larger HT")
#for (i in 1:ncol(mplot))
#  mtext(colnames(mplot)[i],col=colo[i],line=-i)

@@

file = "./Figures/mssa_msse_now.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7, height = 7)

hp_c_out<-filter(x_mat[,select],hp_c,side=1)
par(mfrow=c(1,1))
#mplot<-scale(cbind(target_shifted_mat,hp_c_out,mssa[,select],mssa_HT_HP_C[,select],mmse[,select]))
#dim(mplot)
#colnames(mplot)<-c("Target: HP-BIP","HP-C","M-SSA smaller HT","M-SSA larger HT","M-MSE")
mplot<-scale(cbind(target_shifted_mat,mssa[,select],mmse[,select]))
dim(mplot)
colnames(mplot)<-c("Target: HP-GDP","M-SSA","M-MSE")

rownames(mplot)<-rownames(x_mat)
mplot<-mplot[L:nrow(mplot),]
colo<-c("black","violet","blue","orange","green")
colo<-c("black","blue","green")
main_title<-"Nowcasting HP-GDP"
plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],lwd=1,ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
mtext(colnames(mplot)[1],col=colo[1],line=-1)
for (i in 1:ncol(mplot))
{
  lines(mplot[,i],col=colo[i],lwd=1,lty=1)
  mtext(colnames(mplot)[i],col=colo[i],line=-i)
}
abline(h=0)
abline(v=which(rownames(mplot)>date_to_fit)[1],lty=2,lwd=2)
axis(1,at=c(1,4*1:(nrow(mplot)/4)),labels=rownames(mplot)[c(1,4*1:(nrow(mplot)/4))])
axis(2)
box()

@@

mat_cor_ht<-rbind(c(cor(na.exclude(mplot[,c("Target: HP-GDP","M-SSA")]))[1,2],
cor(na.exclude(mplot[,c("Target: HP-GDP","M-MSE")]))[1,2]),
c(compute_empirical_ht_func(na.exclude(mssa[,select]))$empirical_ht,
compute_empirical_ht_func(na.exclude(mmse[,select]))$empirical_ht))

@@

Figure \ref{bk_gammak} displays the coefficients of the multivariate nowcasts, where increased smoothing achieved through the M-SSA method manifests as a distinct pattern in the filter weights, notably influencing the rate of decay across increasing lags. The resulting filtered series, presented in Figure \ref{mssa_msse_now}, are standardized to enable direct comparisons. The M-SSA nowcast adheres to the specified constraint, exhibiting approximately one-third fewer zero-crossings than the M-MSE, confirming the effectiveness of the smoothness control imposed during the design. This is further corroborated by Table \ref{corhtnow}, which reports empirical holding times and target correlations. Increased smoothness achieved by the M-SSA comes at the cost of a slight reduction in tracking accuracy, as indicated by a marginally lower target correlation. This highlights the inherent trade-off—often referred to as the prediction dilemma—between smoothness (zero-crossing rate) and predictive accuracy (target correlation) within the M-SSA criterion.
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/bk_gammak.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=3in, width=5in]{", file, "}\n",sep = "")
cat("\\caption{Multivariate M-MSE (left) and M-SSA  (right) nowcast filters.", sep = "")
cat("\\label{bk_gammak}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/mssa_msse_now.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=4in, width=5.5in]{", file, "}\n",sep = "")
cat("\\caption{Two-sided HP(160) applied to GDP (target, black) and nowcasts:  M-SSA (blue) and  M-MSE (green). The dashed vertical line delimits in- and out-of-sample spans. The two-sided filter does not extend to the sample end.", sep = "")
cat("\\label{mssa_msse_now}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@
<<label=ats_mba_2,echo=FALSE,results=tex>>=
mat1<-round(mat_cor_ht,3)
#latex(cor_vec, dec = 1, , caption = "Example of using latex to create table",
#center = "centering", file = "", floating = FALSE)
xtable(mat1, dec = 1,digits=rep(2,dim(mat1)[2]+1),
paste("Sample target correlations and HTs of M-SSA and M-MSE nowcasts.  "),
label=paste("corhtnow",sep=""),
center = "centering", file = "", floating = FALSE)
@
The results validate that the smoothing constraint incorporated into the multivariate extension of Criterion \eqref{critssa} effectively manages the zero-crossings of the predictor: sample estimates of the holding times (HTs) align with the expected values derived from Equation \eqref{ht}\footnote{Effective convergence of sample estimates to their theoretical expected values —assuming the true data-generating process is known— can be validated using the M-SSA package. This validation is typically performed through analyses of very long samples of simulated (multivariate) data, ensuring the consistency of the estimation procedures under idealized conditions.}. An increase of $50\%$ in the HT relative to the M-MSE predictor has only a modest effect on the target correlation, suggesting that this tradeoff could be beneficial in our application. Going forward, we now omit the M-MSE, as it is a special case of the more general M-SSA framework.

@@

file = "./Figures/bk_h.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 8, height = 6)

par(mfrow=c(1,2))
# h=0, see above
mplot<-bk
colo<-rainbow(ncol(mplot))
ts.plot(mplot,col=colo,main="M-SSA: h=0",xlab="")
for (i in 1:ncol(mplot))
  mtext(colnames(mplot)[i],col=colo[i],line=-i)

# h=4
bk<-matrix(mssa_indicator_obj$bk_x_array[select,select,,"h=4"],nrow=L)
colnames(bk)<-colnames(gammak)<-select_vec_multi
mplot<-bk
colo<-rainbow(ncol(mplot))
ts.plot(mplot,col=colo,main="M-SSA: h=4",xlab="")
#for (i in 1:ncol(mplot))
#  mtext(colnames(mplot)[i],col=colo[i],line=-i)

@@

file = "./Figures/multivar_vs_univar.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 8, height = 8)

# Select BIP
i<-1
# HP-C Financial crisis
mplot<-scale(hp_c_array[i,,])
colnames(mplot)<-paste("h=",h_vec,sep="")
par(mfrow=c(2,2))
colo<-c(rainbow(ncol(mplot)))
mplot<-mplot[which(rownames(mplot)>2007&rownames(mplot)<2011),]
main_title<-paste("HP-C targeting HP-",colnames(x_mat)[i],sep="")
main_title<-"HP-C: financial crisis"
plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
mtext(colnames(mplot)[1],col=colo[1],line=-1)
for (j in 1:ncol(mplot))
{
  lines(mplot[,j],col=colo[j],lwd=1,lty=1)
  mtext(colnames(mplot)[j],col=colo[j],line=-j)
}
abline(h=0)
axis(1,at=1:nrow(mplot),labels=rownames(mplot))
axis(2)
box()

# HP-C Pandemic
mplot<-scale(hp_c_array[i,,])
colnames(mplot)<-paste("HP-",colnames(x_mat)[i],": h=",h_vec,sep="")
colo<-c(rainbow(ncol(mplot)))
mplot<-mplot[which(rownames(mplot)>2017&rownames(mplot)<2022),]
main_title<-paste("HP-C targeting HP-",colnames(x_mat)[i],sep="")
main_title<-"HP-C: pandemic"
plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
#mtext(colnames(mplot)[1],col=colo[1],line=-1)
for (j in 1:ncol(mplot))
{
  lines(mplot[,j],col=colo[j],lwd=1,lty=1)
#  mtext(colnames(mplot)[j],col=colo[j],line=-j)
}
abline(h=0)
axis(1,at=1:nrow(mplot),labels=rownames(mplot))
axis(2)
box()

@@

colo<-c(rainbow(ncol(mplot)))
main_title<-paste("M-SSA targeting HP-",colnames(x_mat)[i],sep="")
main_title<-"MSSA: financial crisis"
plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
#mtext(colnames(mplot)[1],col=colo[1],line=-1)
for (j in 1:ncol(mplot))
{
  lines(mplot[,j],col=colo[j],lwd=1,lty=1)
#  mtext(colnames(mplot)[j],col=colo[j],line=-j)
}
abline(h=0)
axis(1,at=1:nrow(mplot),labels=rownames(mplot))
axis(2)
box()

@@

# Plot of M-SSA and HP-C, only financial crisis
file = "./Figures/mssa_hpc_financial_crisis.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 8, height = 8)

@@

file = "./Figures/revisions3.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7, height = 7)
par(mfrow=c(1,1))
